---
title: "Agent-Based Modeling in AgenticWaves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Agent-Based Modeling in AgenticWaves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Agent-Based Modeling in AgenticWaves

AgenticWaves implements a sophisticated agent-based modeling (ABM) framework for financial market simulation. This guide explores the ABM components, agent types, and advanced simulation capabilities.

## Overview

The ABM framework in AgenticWaves features:

- **6 distinct behavioral agent types** with realistic trading patterns
- **Heterogeneous agent populations** with customizable diversity
- **Multi-layer network interactions** (trading, information, social)
- **Dynamic wealth distribution** with inequality tracking
- **Crisis-dependent behavior adaptation**
- **Realistic market microstructure** with price impact and transaction costs

## Agent Types

### 1. Momentum Traders
Follow price trends and market momentum:

```{r}
# Momentum traders characteristics
- High trend sensitivity (0.9)
- Moderate risk tolerance (0.7)
- High trading frequency (0.8)
- Short memory length (20 periods)
- Strategy: Buy rising assets, sell falling assets
```

### 2. Contrarian Traders
Trade against prevailing market trends:

```{r}
# Contrarian traders characteristics
- Negative trend sensitivity (-0.7)
- Moderate risk tolerance (0.6)
- Moderate trading frequency (0.5)
- Long memory length (50 periods)
- Strategy: Buy oversold assets, sell overbought assets
```

### 3. Fundamentalist Traders
Base decisions on fundamental analysis:

```{r}
# Fundamentalist traders characteristics
- Low trend sensitivity (0.2)
- Conservative risk tolerance (0.5)
- Low trading frequency (0.3)
- Very long memory (100 periods)
- Strategy: Mean reversion toward fundamental value
```

### 4. Noise Traders
Make random or irrational trading decisions:

```{r}
# Noise traders characteristics
- Very low trend sensitivity (0.1)
- High risk tolerance (0.9)
- Very high trading frequency (0.9)
- Very short memory (5 periods)
- Strategy: Random trading decisions
```

### 5. Herding Traders
Follow crowd behavior and social signals:

```{r}
# Herding traders characteristics
- Moderate trend sensitivity (0.5)
- Moderate risk tolerance (0.6)
- Moderate trading frequency (0.6)
- Short memory length (15 periods)
- Strategy: Follow dominant market sentiment
```

### 6. Sophisticated Traders
Use complex multi-factor strategies:

```{r}
# Sophisticated traders characteristics
- High trend sensitivity (0.6)
- Conservative risk tolerance (0.4)
- Low trading frequency (0.4)
- Very long memory (200 periods)
- Strategy: Multi-factor analysis with risk management
```

## Creating Agent Populations

### Basic Population Creation

```{r}
library(AgenticWaves)

# Create a diverse agent population
agents <- create_enhanced_agent_population(
  n_agents = 500,
  behavioral_heterogeneity = 0.7,
  wealth_distribution = "pareto"
)

# View population summary
print(attr(agents, "type_distribution"))
print(paste("Initial wealth Gini:", round(attr(agents, "wealth_gini"), 3)))
```

### Wealth Distribution Options

```{r}
# Equal wealth distribution
agents_equal <- create_enhanced_agent_population(
  n_agents = 200,
  wealth_distribution = "equal"
)

# Normal wealth distribution
agents_normal <- create_enhanced_agent_population(
  n_agents = 200,
  wealth_distribution = "normal"
)

# Pareto wealth distribution (realistic inequality)
agents_pareto <- create_enhanced_agent_population(
  n_agents = 200,
  wealth_distribution = "pareto"
)
```

### Behavioral Heterogeneity

The `behavioral_heterogeneity` parameter controls diversity within agent types:

```{r}
# Low heterogeneity (similar agents within types)
agents_low_het <- create_enhanced_agent_population(
  n_agents = 300,
  behavioral_heterogeneity = 0.3
)

# High heterogeneity (diverse agents within types)
agents_high_het <- create_enhanced_agent_population(
  n_agents = 300,
  behavioral_heterogeneity = 0.9
)
```

## Multi-Layer Networks

### Network Types

AgenticWaves creates three types of agent networks:

```{r}
# Create multi-layer network
network <- create_dynamic_multilayer_network(
  agents = agents,
  network_types = c("trading", "information", "social"),
  density = 0.1
)

# Access individual networks
trading_network <- network$networks$trading
information_network <- network$networks$information
social_network <- network$networks$social

# View network properties
print(trading_network$metrics)
```

### Trading Network
Connections based on:
- Similar agent types
- Similar wealth levels  
- Similar risk tolerance

### Information Network
Connections based on:
- Agent sophistication
- Memory length
- Information processing capabilities

### Social Network
Small-world network with:
- Lattice structure with random rewiring
- Local clustering with long-range connections
- Realistic social interaction patterns

## Market Simulation

### Basic Simulation

```{r}
# Load sample data
data <- get_sample_data("global_markets", n_assets = 10, n_periods = 500)

# Run market simulation
sim_results <- simulate_enhanced_market_dynamics(
  agents = agents,
  asset_data = data,
  n_periods = 1000,
  network_effects = TRUE
)
```

### Simulation Components

The simulation includes:

1. **Market Regime Dynamics**: Volatility regimes switch based on market conditions
2. **Agent Decision Making**: Each agent makes trading decisions based on their type
3. **Network Information Flow**: Information spreads through agent networks
4. **Price Impact**: Agent trades affect market prices
5. **Transaction Costs**: Realistic trading costs reduce profits
6. **Wealth Evolution**: Agent wealth changes based on trading performance

### Market Regimes

```{r}
# Market regimes in simulation
# 1 = Low volatility
# 2 = Normal volatility  
# 3 = High volatility/crisis

regime_summary <- table(sim_results$market_regimes)
print(regime_summary)

# Crisis probability and transitions
crisis_periods <- which(sim_results$market_regimes == 3)
print(paste("Crisis periods:", length(crisis_periods)))
```

### Agent Performance Analysis

```{r}
# Extract agent types
agent_types <- sapply(agents, function(x) x$type)

# Calculate performance by type
performance_by_type <- tapply(sim_results$agent_returns, agent_types, function(x) {
  data.frame(
    mean_return = mean(x),
    volatility = sd(x),
    sharpe_ratio = mean(x) / sd(x),
    max_return = max(x),
    min_return = min(x)
  )
})

# View performance summary
print(performance_by_type)
```

### Wealth Inequality Evolution

```{r}
# Track wealth inequality over time
wealth_gini_evolution <- sim_results$wealth_evolution_gini

# Plot wealth inequality
plot(wealth_gini_evolution, 
     type = "l", 
     main = "Wealth Inequality Evolution",
     xlab = "Time Period", 
     ylab = "Gini Coefficient")

# Final wealth distribution
final_wealth <- sim_results$agent_wealth[nrow(sim_results$agent_wealth), ]
final_gini <- calculate_gini_coefficient(final_wealth)

print(paste("Final wealth Gini:", round(final_gini, 3)))
```

## Advanced Agent Behaviors

### Crisis Adaptation

Agents adapt their behavior during crisis periods:

```{r}
# During crisis (regime 3):
# - Risk tolerance decreases
# - Trading frequency may increase (flight to quality)
# - Herding behavior intensifies
# - Fundamental analysis becomes less reliable
```

### Social Learning

Agents learn from connected neighbors:

```{r}
# Information sharing through networks:
# - Sophisticated agents share more information
# - Information quality depends on agent type
# - Social influence affects trading decisions
# - Network structure impacts information flow
```

### Memory and Learning

Each agent maintains memory of past market conditions:

```{r}
# Memory effects:
# - Recent events weighted more heavily
# - Memory length varies by agent type
# - Learning rate affects adaptation speed
# - Performance feedback influences future decisions
```

## Customizing Agent Behavior

### Creating Custom Agent Types

While AgenticWaves provides 6 built-in agent types, you can customize behavior:

```{r}
# Modify agent parameters after creation
custom_agents <- create_enhanced_agent_population(n_agents = 100)

# Customize specific agents
for(i in 1:length(custom_agents)) {
  if(custom_agents[[i]]$type == "momentum") {
    # Make momentum traders more aggressive
    custom_agents[[i]]$risk_tolerance <- custom_agents[[i]]$risk_tolerance * 1.2
    custom_agents[[i]]$trading_frequency <- custom_agents[[i]]$trading_frequency * 1.1
  }
}
```

### Parameter Sensitivity Analysis

Test how different parameters affect outcomes:

```{r}
# Test different behavioral heterogeneity levels
heterogeneity_levels <- c(0.3, 0.5, 0.7, 0.9)
results_by_heterogeneity <- list()

for(het in heterogeneity_levels) {
  agents_test <- create_enhanced_agent_population(
    n_agents = 200,
    behavioral_heterogeneity = het
  )
  
  sim_test <- simulate_enhanced_market_dynamics(
    agents = agents_test,
    asset_data = data,
    n_periods = 500
  )
  
  results_by_heterogeneity[[as.character(het)]] <- sim_test$final_wealth_gini
}

print(results_by_heterogeneity)
```

## Visualization of Agent Networks

### Agent Network by Type

```{r}
# Create agent network visualization
network_plot <- plot_agent_type_network(
  agents = agents,
  network_obj = network$networks$trading$graph,
  layout = "stress"
)

print(network_plot)
```

### Wealth Distribution Plots

```{r}
# Plot wealth distribution by agent type
wealth_by_type <- data.frame(
  agent_type = sapply(agents, function(x) x$type),
  wealth = sapply(agents, function(x) x$current_wealth)
)

# Wealth distribution plot
library(ggplot2)
ggplot(wealth_by_type, aes(x = agent_type, y = wealth, fill = agent_type)) +
  geom_boxplot() +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Wealth Distribution by Agent Type",
       x = "Agent Type", y = "Wealth (log scale)")
```

## Performance Optimization

### Simulation Speed

For large simulations, consider:

```{r}
# Optimize for speed
# - Reduce number of agents for initial testing
# - Use shorter simulation periods
# - Disable network effects for simple simulations
# - Use parallel processing where available

# Fast simulation for testing
quick_sim <- simulate_enhanced_market_dynamics(
  agents = agents[1:100],  # Subset of agents
  asset_data = data,
  n_periods = 250,         # Shorter period
  network_effects = FALSE  # Disable networks
)
```

### Memory Management

```{r}
# For large simulations:
# - Monitor memory usage
# - Clear intermediate results
# - Use appropriate data types
# - Consider chunked processing for very long simulations

# Monitor memory (if pryr package available)
if(requireNamespace("pryr", quietly = TRUE)) {
  print(pryr::mem_used())
}
```

## Research Applications

The ABM framework supports various research questions:

### Market Efficiency Studies
- Impact of different trader types on price discovery
- Role of information networks in market efficiency
- Behavioral biases and market anomalies

### Systemic Risk Analysis  
- Contagion mechanisms through agent networks
- Impact of wealth inequality on market stability
- Crisis propagation and amplification

### Policy Analysis
- Effects of transaction taxes on different agent types
- Impact of market structure changes
- Regulatory interventions and market outcomes

### Behavioral Finance
- Herding behavior and bubble formation
- Momentum vs. contrarian strategies
- Learning and adaptation in financial markets

## Next Steps

- Explore **Network Analysis** for spillover detection
- Read **Visualization Guide** for advanced plotting
- Try **Advanced Analysis** for complex workflows
- Check **Custom Data** tutorial for your own datasets

## References

1. Axtell, R. L., & Farmer, J. D. (2025). Agent-based modeling in economics and finance: Past, present, and future.
2. Hommes, C. (2006). Heterogeneous agent models in economics and finance.
3. LeBaron, B. (2006). Agent-based computational finance.
4. Farmer, J. D., & Foley, D. (2009). The economy needs agent-based modelling.