---
title: "Network Analysis and Spillover Detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network Analysis and Spillover Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Network Analysis and Spillover Detection

AgenticWaves provides advanced network analysis capabilities for detecting spillover effects, contagion episodes, and systemic risk in financial markets. This guide covers the comprehensive network analysis framework.

## Overview

The network analysis framework includes:

- **Dynamic spillover networks** with time-varying connectivity
- **Multi-methodology contagion detection** using consensus algorithms
- **Regime-switching network dynamics** with structural break detection
- **Comprehensive network metrics** and centrality measures
- **Real-time spillover monitoring** with significance testing
- **Publication-quality network visualizations**

## Core Concepts

### Spillover Effects
Spillover effects represent the transmission of shocks between financial assets or markets:

- **Return spillovers**: Price movement transmission
- **Volatility spillovers**: Risk transmission  
- **Correlation spillovers**: Co-movement changes
- **Liquidity spillovers**: Funding constraint transmission

### Network Construction
Networks are built from statistical relationships:

```{r}
# Common network construction methods:
# 1. Correlation networks (linear relationships)
# 2. Partial correlation networks (direct relationships)
# 3. Transfer entropy networks (information flow)
# 4. Granger causality networks (predictive relationships)
```

## Dynamic Spillover Analysis

### Basic Spillover Calculation

```{r}
library(AgenticWaves)

# Load data and run simulation
data <- get_sample_data("global_markets", n_assets = 10, n_periods = 500)
agents <- create_enhanced_agent_population(n_agents = 300)

sim_results <- simulate_enhanced_market_dynamics(
  agents = agents,
  asset_data = data,
  n_periods = 1000
)

# Calculate dynamic spillovers
spillover_results <- calculate_dynamic_spillover_networks(
  simulation_results = sim_results,
  window_size = 100,
  significance_level = 0.05
)
```

### Spillover Components

The spillover analysis provides:

```{r}
# Main spillover measures
total_spillover <- spillover_results$total_spillover          # Overall connectivity
directional_spillovers <- spillover_results$directional_spillovers  # Asset-specific spillovers
network_density <- spillover_results$network_density         # Network density evolution
network_clustering <- spillover_results$network_clustering   # Clustering evolution

# Summary statistics
cat("Average total spillover:", round(mean(total_spillover), 2), "%\n")
cat("Peak spillover:", round(max(total_spillover), 2), "%\n")
cat("Spillover volatility:", round(sd(total_spillover), 2), "%\n")
```

### Rolling Window Analysis

The rolling window approach captures dynamic relationships:

```{r}
# Window size considerations:
# - Larger windows: More stable estimates, less responsive
# - Smaller windows: More responsive, noisier estimates
# - Typical range: 50-200 observations

# Test different window sizes
window_sizes <- c(50, 100, 150, 200)
spillover_by_window <- list()

for(ws in window_sizes) {
  spillover_temp <- calculate_dynamic_spillover_networks(
    sim_results, 
    window_size = ws
  )
  spillover_by_window[[paste0("window_", ws)]] <- spillover_temp$total_spillover
}

# Compare spillover evolution
par(mfrow = c(2, 2))
for(i in 1:length(window_sizes)) {
  plot(spillover_by_window[[i]], 
       type = "l", 
       main = paste("Window Size:", window_sizes[i]),
       ylab = "Total Spillover (%)")
}
```

## Advanced Contagion Detection

### Multi-Methodology Detection

```{r}
# Use multiple detection methods for robust results
contagion_results <- detect_contagion_episodes(
  spillover_results = spillover_results,
  detection_methods = c("threshold", "regime", "correlation", "volatility")
)

# View consensus episodes
consensus_episodes <- contagion_results$consensus_episodes
episode_characteristics <- contagion_results$episode_characteristics

print(episode_characteristics)
```

### Detection Methods

#### 1. Threshold-Based Detection
Identifies periods with unusually high spillovers:

```{r}
# Multiple threshold levels
thresholds <- c(0.75, 0.85, 0.90, 0.95)  # Quantile thresholds

# Episodes at different threshold levels
threshold_episodes <- contagion_results$individual_indicators$threshold
```

#### 2. Regime-Switching Detection
Uses changepoint analysis to identify regime shifts:

```{r}
# Structural breaks in spillover evolution
regime_episodes <- contagion_results$individual_indicators$regime

# Regime characteristics:
# - High spillover regimes vs. low spillover regimes
# - Regime duration and intensity
# - Transition probabilities
```

#### 3. Correlation-Based Detection
Identifies periods of unusually high cross-asset correlations:

```{r}
# High correlation episodes
correlation_episodes <- contagion_results$individual_indicators$correlation

# Features:
# - Average correlation evolution
# - Correlation spikes and clusters
# - Flight-to-quality episodes
```

#### 4. Volatility-Based Detection
Detects volatility spikes that often accompany contagion:

```{r}
# Volatility spike detection
volatility_episodes <- contagion_results$individual_indicators$volatility

# Characteristics:
# - Synchronized volatility increases
# - Volatility clustering
# - Risk-off periods
```

## Network Metrics and Analysis

### Comprehensive Network Metrics

```{r}
# Calculate detailed network metrics
network_metrics <- calculate_network_metrics(
  spillover_results$network_results$graph
)

# Basic properties
print(paste("Number of nodes:", network_metrics$n_nodes))
print(paste("Number of edges:", network_metrics$n_edges))
print(paste("Network density:", round(network_metrics$density, 3)))
print(paste("Average clustering:", round(network_metrics$transitivity, 3)))
```

### Centrality Measures

Different centrality measures reveal different aspects of network importance:

```{r}
# Degree centrality (number of connections)
degree_centrality <- network_metrics$degree_centrality

# Betweenness centrality (bridge nodes)
betweenness_centrality <- network_metrics$betweenness_centrality

# Closeness centrality (information spread)
closeness_centrality <- network_metrics$closeness_centrality

# Eigenvector centrality (influence)
eigenvector_centrality <- network_metrics$eigenvector_centrality

# Identify most central nodes
most_central_degree <- which.max(degree_centrality$values)
most_central_betweenness <- which.max(betweenness_centrality$values)

cat("Most connected asset (degree):", colnames(data)[most_central_degree], "\n")
cat("Most important bridge (betweenness):", colnames(data)[most_central_betweenness], "\n")
```

### Community Detection

Identify clusters of highly connected assets:

```{r}
# Community structure
community_info <- network_metrics$community_membership
n_communities <- network_metrics$n_communities
modularity_score <- network_metrics$modularity

cat("Number of communities:", n_communities, "\n")
cat("Modularity score:", round(modularity_score, 3), "\n")

# Assets by community
for(i in 1:n_communities) {
  community_assets <- which(community_info == i)
  cat("Community", i, ":", colnames(data)[community_assets], "\n")
}
```

### Small-World Properties

Test for small-world network characteristics:

```{r}
# Small-world properties
small_world_sigma <- network_metrics$small_world_sigma

if(!is.na(small_world_sigma)) {
  cat("Small-world sigma:", round(small_world_sigma, 3), "\n")
  
  if(small_world_sigma > 1) {
    cat("Network exhibits small-world properties\n")
  } else {
    cat("Network does not exhibit small-world properties\n")
  }
}
```

## Time-Varying Network Analysis

### Network Evolution

Track how network structure changes over time:

```{r}
# Network metrics evolution
plot(spillover_results$network_density, 
     type = "l", 
     main = "Network Density Evolution",
     xlab = "Time Window", 
     ylab = "Density")

plot(spillover_results$network_clustering, 
     type = "l", 
     main = "Network Clustering Evolution",
     xlab = "Time Window", 
     ylab = "Clustering Coefficient")
```

### Structural Break Detection

Identify periods of significant network structure changes:

```{r}
# Structural breaks in network properties
structural_breaks <- spillover_results$structural_breaks

if(length(structural_breaks$density) > 0) {
  cat("Density breaks at windows:", structural_breaks$density, "\n")
}

if(length(structural_breaks$clustering) > 0) {
  cat("Clustering breaks at windows:", structural_breaks$clustering, "\n")
}
```

## Asset-Level Spillover Analysis

### Directional Spillovers

Analyze which assets are spillover transmitters vs. receivers:

```{r}
# Asset spillover statistics
asset_stats <- spillover_results$asset_spillover_stats

# Net spillover contributors (positive = transmitter, negative = receiver)
net_spillovers <- asset_stats$net_spillover
names(net_spillovers) <- colnames(data)

# Top spillover transmitters
transmitters <- sort(net_spillovers, decreasing = TRUE)[1:3]
cat("Top spillover transmitters:\n")
print(transmitters)

# Top spillover receivers  
receivers <- sort(net_spillovers, decreasing = FALSE)[1:3]
cat("Top spillover receivers:\n")
print(receivers)
```

### Spillover Volatility

Identify assets with variable spillover roles:

```{r}
# Assets with high spillover volatility
spillover_volatility <- asset_stats$spillover_volatility
names(spillover_volatility) <- colnames(data)

variable_spillovers <- sort(spillover_volatility, decreasing = TRUE)[1:3]
cat("Most variable spillover roles:\n")
print(variable_spillovers)
```

## Crisis vs. Normal Period Analysis

### Spillover Comparison

Compare spillover patterns during different market conditions:

```{r}
# Identify crisis periods (high volatility regime)
crisis_periods <- which(sim_results$market_regimes == 3)
normal_periods <- which(sim_results$market_regimes == 1)

# Map to spillover windows (accounting for window lag)
crisis_windows <- pmax(1, crisis_periods - spillover_results$window_size)
normal_windows <- pmax(1, normal_periods - spillover_results$window_size)

# Crisis vs normal spillovers
crisis_spillovers <- total_spillover[crisis_windows]
normal_spillovers <- total_spillover[normal_windows]

cat("Average crisis spillover:", round(mean(crisis_spillovers, na.rm = TRUE), 2), "%\n")
cat("Average normal spillover:", round(mean(normal_spillovers, na.rm = TRUE), 2), "%\n")
cat("Crisis spillover increase:", 
    round(mean(crisis_spillovers, na.rm = TRUE) / mean(normal_spillovers, na.rm = TRUE) - 1, 2) * 100, "%\n")
```

### Network Structure Changes

Compare network properties across market regimes:

```{r}
# Network density by regime
crisis_density <- network_density[crisis_windows]
normal_density <- network_density[normal_windows]

cat("Crisis network density:", round(mean(crisis_density, na.rm = TRUE), 3), "\n")
cat("Normal network density:", round(mean(normal_density, na.rm = TRUE), 3), "\n")

# Clustering by regime
crisis_clustering <- network_clustering[crisis_windows]
normal_clustering <- network_clustering[normal_windows]

cat("Crisis clustering:", round(mean(crisis_clustering, na.rm = TRUE), 3), "\n")
cat("Normal clustering:", round(mean(normal_clustering, na.rm = TRUE), 3), "\n")
```

## Advanced Visualization

### Dynamic Network Plots

Create time-varying network visualizations:

```{r}
# Network plot for specific time window
window_index <- 50  # Choose window

# Extract network for this window
window_spillovers <- matrix(spillover_results$spillover_evolution[window_index, ], 
                           nrow = sqrt(ncol(spillover_results$spillover_evolution)))

# Create network plot
network_plot <- plot_enhanced_network(
  window_spillovers,
  layout = "stress",
  node_size_var = "degree",
  color_scheme = "viridis"
)

print(network_plot)
```

### Spillover Heatmaps

Visualize spillover matrices:

```{r}
# Create spillover heatmap
library(ggplot2)

# Average spillover matrix
avg_spillover_matrix <- matrix(
  colMeans(spillover_results$spillover_evolution),
  nrow = ncol(data)
)

# Convert to long format for ggplot
heatmap_data <- expand.grid(
  From = colnames(data),
  To = colnames(data)
)
heatmap_data$Spillover <- as.vector(avg_spillover_matrix)

# Create heatmap
ggplot(heatmap_data, aes(x = From, y = To, fill = Spillover)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Spillover\nIntensity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Average Spillover Matrix",
       x = "Spillover From", y = "Spillover To")
```

## Statistical Significance Testing

### Spillover Significance

Test statistical significance of spillover measures:

```{r}
# Significance testing built into calculation
significant_spillovers <- spillover_results$spillover_evolution
significant_spillovers[significant_spillovers == 0] <- NA  # Non-significant set to NA

# Proportion of significant spillovers over time
significance_ratio <- apply(spillover_results$spillover_evolution > 0, 1, mean)

plot(significance_ratio, 
     type = "l",
     main = "Proportion of Significant Spillovers",
     xlab = "Time Window",
     ylab = "Significance Ratio")
```

### Bootstrap Confidence Intervals

Generate confidence intervals for spillover estimates:

```{r}
# Bootstrap procedure (simplified example)
bootstrap_spillovers <- function(data, n_bootstrap = 100) {
  bootstrap_results <- matrix(0, n_bootstrap, length(total_spillover))
  
  for(b in 1:n_bootstrap) {
    # Resample data
    boot_indices <- sample(nrow(data), replace = TRUE)
    boot_data <- data[boot_indices, ]
    
    # Recalculate spillovers (simplified)
    # In practice, would run full simulation
    boot_spillovers <- rnorm(length(total_spillover), 
                           mean = mean(total_spillover), 
                           sd = sd(total_spillover))
    bootstrap_results[b, ] <- boot_spillovers
  }
  
  return(bootstrap_results)
}

# Generate confidence intervals
# boot_results <- bootstrap_spillovers(data)
# ci_lower <- apply(boot_results, 2, quantile, 0.025)
# ci_upper <- apply(boot_results, 2, quantile, 0.975)
```

## Research Applications

### Systemic Risk Assessment
- Identify systemically important assets
- Monitor system-wide connectivity
- Early warning indicators for financial stress

### Contagion Analysis
- Crisis transmission mechanisms
- Speed and magnitude of contagion
- Policy intervention effectiveness

### Portfolio Management
- Dynamic correlation structures
- Regime-dependent diversification benefits
- Risk management strategies

### Regulatory Analysis
- Market structure impacts
- Interconnectedness measures
- Macroprudential policy design

## Performance Optimization

### Computational Efficiency

For large-scale analysis:

```{r}
# Optimize spillover calculations
# - Use appropriate window sizes
# - Consider parallel processing
# - Optimize significance testing
# - Use efficient network algorithms

# Monitor computation time
system.time({
  spillover_results <- calculate_dynamic_spillover_networks(
    sim_results,
    window_size = 100
  )
})
```

### Memory Management

```{r}
# For large networks:
# - Sparse matrix representations
# - Chunked processing
# - Efficient data structures
# - Memory monitoring

# Clear large objects when done
# rm(large_spillover_object)
# gc()  # Garbage collection
```

## Next Steps

- Explore **Visualization Guide** for advanced network plots
- Read **Advanced Analysis** for complex multi-asset workflows  
- Try **Custom Data** tutorial with your own financial data
- Check **Agent-Based Modeling** for simulation details

## References

1. Baruník, J., & Křehlík, T. (2018). Measuring the frequency dynamics of financial connectedness and systemic risk.
2. Diebold, F. X., & Yılmaz, K. (2012). Better to give than to receive: Predictive directional measurement of volatility spillovers.
3. Billio, M., et al. (2012). Econometric measures of connectedness and systemic risk in the finance and insurance sectors.
4. Acemoglu, D., et al. (2012). The network origins of aggregate fluctuations.